---
title: "Summative Assessment 1"
author: "PATAYON, SPIKE LEE-ROY V"
date: "2025-03-16"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summative Assessment 1

Objective: The purpose of this assessment is to evaluate your understanding of exploratory data analysis techniques, including univariate, bivariate, and trivariate/hypervariate data exploration using computational tools and visualizations.

Dataset: EDA_Ecommerce_Assessment.csv
Dataset Description: The dataset contains information about customer purchasing behavior in an e-commerce platform. The variables include:

* Customer_ID: Unique identifier for each customer
* Gender: Male or Female
* Age: Customer's age in years
* Browsing_Time: Average time spent on the website per visit (in minutes)
* Purchase_Amount: Total amount spent in a single transaction (in USD)
* Number_of_Items: Number of items purchased per transaction
* Discount_Applied: Discount percentage applied to the transaction
* Total_Transactions: Total number of transactions by the customer
* Category: Product category (e.g., Electronics, Clothing, Home & Kitchen, etc.)
* Satisfaction_Score: Customer satisfaction score (1-5 scale)

## Unit 1: Univariate Data Analysis
1. Load the dataset and summarize its structure.
2. Create histograms and boxplots to visualize the distribution of Purchase_Amount, Number_of_Items, and Satisfaction_Score.
3. Compute measures of central tendency (mean, median, mode) and spread (variance, standard deviation, IQR) for Purchase_Amount.
4. Compare the distribution of Browsing_Time and Purchase_Amount across different Gender groups using density plots.
5. Apply a logarithmic or square root transformation on Browsing_Time and evaluate changes in skewness.
6. Fit a simple linear regression model predicting Purchase_Amount based on Browsing_Time. Interpret the results.
7. Use ggplot2 (or equivalent) to create scatter plots and regression lines.

### Part 1:
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(e1071)
```

Load the data set:
```{r}
data <- read.csv("C:\\Users\\spike\\Downloads\\EDA_Ecommerce_Assessment.csv")
```
```{r}
head(data)
```
Here is a quick summarization of each elements in the csv file:
```{r}
summary(data)
```
### Part 2:
Histogram of Purchase_Amount, Number_of_items, and Satisfaction_Score
```{r}
ggplot(data, aes(x = Purchase_Amount)) + 
  geom_histogram( fill = "blue", alpha = 1, color = "black") + 
  ggtitle("Histogram of Purchase Amount")

ggplot(data, aes(x = Number_of_Items )) + 
  geom_histogram( fill = "navy", alpha = 1, color = "black") + 
  ggtitle("Histogram of Purchase Amount")

ggplot(data, aes(x = Satisfaction_Score )) + 
  geom_histogram( fill = "royalblue", alpha = 1, color = "black") + 
  ggtitle("Histogram of Purchase Amount")
```

Boxplot of Purchase_Amount, Number_of_items, and Satisfaction_Score
```{r}
ggplot(data, aes(y = Purchase_Amount)) + 
  geom_boxplot(fill = "orange4", alpha = 1, color = "black") + 
  ggtitle("Boxplot of Purchase Amount")

ggplot(data, aes(y = Number_of_Items)) + 
  geom_boxplot(fill = "orange3", alpha = 1, color = "black") + 
  ggtitle("Boxplot of Number of Items")

ggplot(data, aes(y = Satisfaction_Score)) + 
  geom_boxplot(fill = "orange2", alpha = 1, color = "black") + 
  ggtitle("Boxplot of Satisfaction Score")
```

### Part 3
central tendency (mean, median, mode) and spread (variance, standard deviation, IQR) of Purchase_Amount
```{r}
# Central Tendency
mean_value <- mean(data$Purchase_Amount, na.rm = TRUE)
median_value <- median(data$Purchase_Amount, na.rm = TRUE)
mode_value <- as.numeric(names(sort(table(data$Purchase_Amount), decreasing = TRUE)[1])) 

# Spread
variance_value <- var(data$Purchase_Amount, na.rm = TRUE)
sd_value <- sd(data$Purchase_Amount, na.rm = TRUE)
iqr_value <- IQR(data$Purchase_Amount, na.rm = TRUE)

list(Mean = mean_value, Median = median_value, Mode = mode_value, 
     Variance = variance_value, SD = sd_value, IQR = iqr_value)
```

### Part 4
```{r}
ggplot(data, aes(x = Browsing_Time, fill = Gender)) +
  geom_density(alpha = 0.5) +
  ggtitle("Density Plot of Browsing Time by Gender")

ggplot(data, aes(x = Purchase_Amount, fill = Gender)) +
  geom_density(alpha = 0.5) +
  ggtitle("Density Plot of Purchase Amount by Gender")
```
### Part 5
```{r}
# Log Transformation
data$Browsing_Time_log <- log1p(data$Browsing_Time)
log_skewness <- skewness(data$Browsing_Time_log, na.rm = TRUE)

# Square Root Transformation
data$Browsing_Time_sqrt <- sqrt(data$Browsing_Time)
sqrt_skewness <- skewness(data$Browsing_Time_sqrt, na.rm = TRUE)

# Print Skewness
list(Log_Skewness = log_skewness, Sqrt_Skewness = sqrt_skewness)
```
### Part 6
```{r}
lm_model <- lm(Purchase_Amount ~ Browsing_Time, data = data)
summary(lm_model)
```
### Part 7
```{r}
ggplot(data, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  ggtitle("Scatter Plot of Browsing Time vs Purchase Amount with Regression Line")
```
## Unit 2: Bivariate Data Analysis
8. Create scatter plots to explore the relationship between Purchase_Amount and Number_of_Items.
9. Fit a polynomial regression model for Purchase_Amount and Browsing_Time and compare it with a simple linear model.
10. Apply LOESS (Locally Estimated Scatterplot Smoothing) to Purchase_Amount vs. Browsing_Time and visualize the results.
11. Compare robust regression methods (Huber or Tukey regression) with ordinary least squares (OLS).


### Part 8
```{r}
ggplot(data, aes(x = Number_of_Items, y = Purchase_Amount )) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  ggtitle("Scatter Plot of Purchase Amount vs. Number of Items")

```

### Part 9

```{r}
lm_model <- lm(Purchase_Amount ~ Browsing_Time, data = data)
summary(lm_model)
```
```{r}
poly_model <- lm(Purchase_Amount ~ poly(Browsing_Time, 2, raw = TRUE), data = data)
summary(poly_model)
```
```{r}
anova(lm_model, poly_model)
```
remark: If the p-value from anova() is small, the polynomial model provides a significantly better fit than the simple linear model.

to better visualize it:

```{r}
ggplot(data, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE), color = "green") +
  geom_smooth(method = "lm", color = "red") + 
  ggtitle("Polynomial vs. Linear Regression on Purchase Amount")
```

### Part 10

```{r}
ggplot(data, aes(x = Browsing_Time, y = Purchase_Amount)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "purple") +
  ggtitle("LOESS Smoothing: Purchase Amount vs. Browsing Time")

```

### Part 11

```{r}
ols_model <- lm(Purchase_Amount ~ Browsing_Time, data = data)
summary(ols_model)

```

```{r}
library(MASS)
huber_model <- rlm(Purchase_Amount ~ Browsing_Time, data = data)
summary(huber_model)

```
```{r}
library(robustbase)
tukey_model <- lmrob(Purchase_Amount ~ Browsing_Time, data = data)
summary(tukey_model)

```
```{r}
summary(ols_model)$r.squared  # R-squared of OLS
summary(huber_model)$r.squared  # R-squared of Huber
summary(tukey_model)$r.squared  # R-squared of Tukey

```

OLS is sensitive to outliers, while Huber and Tukey regressions handle them better.

Interpretation:

If R-squared is much lower for OLS but remains stable for Huber or Tukey, it suggests that outliers are influencing OLS.
If Tukey’s regression shows improvement, non-Gaussian noise is likely present.

## Unit 3: Trivariate/Hypervariate Data Analysis
12. Explore interaction effects between Browsing_Time and Category on Purchase_Amount using interaction plots. 

13. Create coplots of Purchase_Amount against Browsing_Time for different levels of Category. 

14. Use level plots or contour plots to visualize relationships between Browsing_Time, Number_of_Items, and Purchase_Amount. 

15. Perform multiple regression with Purchase_Amount as the dependent variable and Browsing_Time, Number_of_Items, and Satisfaction_Score as predictors. Perform model selection and assess variable importance.

### Part 12
```{r}
library(interactions)
library(ggplot2)

# Interaction plot using ggplot2
ggplot(data, aes(x = Browsing_Time, y = Purchase_Amount, color = Category)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Interaction Effect of Browsing Time and Category on Purchase Amount")

```

```{r}
ggplot(data, aes(x = Browsing_Time, y = Purchase_Amount, color = Category)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  ggtitle("LOESS Interaction Effect of Browsing Time and Category on Purchase Amount")

```



### Part 13
```{r}
library(lattice)

# Coplot: Purchase Amount vs Browsing Time for different Categories
coplot(Purchase_Amount ~ Browsing_Time | Category, data = data,
       panel = panel.smooth)

```
### Part 14
```{r}
library(ggplot2)

ggplot(data, aes(x = Browsing_Time, y = Number_of_Items, fill = Purchase_Amount)) +
  geom_tile() +
  scale_fill_viridis_c() + 
  ggtitle("Level Plot of Browsing Time, Number of Items, and Purchase Amount")

```

### Part 15

```{r}
multi_model <- lm(Purchase_Amount ~ Browsing_Time + Number_of_Items + Satisfaction_Score, data = data)
summary(multi_model)

```
```{r}
library(MASS)

# Stepwise selection using AIC
stepwise_model <- stepAIC(multi_model, direction = "both")
summary(stepwise_model)

```
```{r}
library(car)

vif(multi_model)  # Variance Inflation Factor (detects multicollinearity)

library(caret)

# Calculate importance
importance <- varImp(multi_model, scale = TRUE)
print(importance)

```

